Challenges to the Two-Infall Scenario by Large Stellar Age Catalogs

### 2025-12-09 src/tex/ms.tex

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{\figpath{stellar_abundance_evolution.pdf}}
    \caption{Stellar age--abundance relations predicted by multi-zone models at the \yZ{1} yield scale (see Table \ref{tab:yields}). Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}< 9\kpc$, $|z| < 0.5\kpc$) color-coded by its birth radius. In this and subsequent figures, a Gaussian scatter is applied to each point according to the median age and abundance uncertainties in Table \ref{tab:uncertainties}. For visual clarity, we plot only a random mass-weighted sample of \num{10000} points in each panel. The dashed curve plots the predicted ISM abundance at $R_{\rm gal}=8\kpc$, and the black squares plot the median stellar abundance in {2 Gyr}-wide age bins. The red line and points are the median abundance for APOGEE stars, and the shaded regions are the 16th--84th percentile ranges. Each column shows results from a different multi-zone model: {\bf (a)} our fiducial model, {\tt yZ1-fiducial}; {\bf (b)} a model with stronger radial migration, {\yZ1-migration}; {\bf (c)} a model with a higher local thick-to-thin disk ratio, {\yZ1-diskratio}; and {\bf (d)} a model with pre-enriched gas infall, {\tt yZ1-preenrich} (see Table \ref{tab:models} for details).}
    \label{fig:abundance-evolution-params}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{\figpath{ofe_feh_density.pdf}}
    \caption{The density of stars in the [O/Fe]--[Fe/H] plane predicted by multi-zone models with (a) $y/Z_\odot=1$ and (b) $y/Z_\odot=2$, and (c) from the APOGEE DR17 catalog. The curves in panels (a) and (b) plot the ISM abundance at the Solar annulus over time, and the alternating black and white segments mark time intervals of {1 Gyr}. The model output has been re-sampled to match the APOGEE stellar $|z|$ distribution.
    Stars in all panels are restricted to the region defined by $7\leq R_{\rm gal}< 9\kpc$ and $|z|<2\kpc$.}
    \label{fig:ofe-feh-density}
\end{figure}

@@

\added{A few recent studies have reported the discovery of multiple disconnected sequences in the local age--metallicity relation \citep[e.g.,][]{sahlholdt_characterizing_2022,xiang_time-resolved_2022,anders_spectroscopic_2023}. \citet{nissen_high-precision_2020} argued that these sequences provide evidence for the two-infall scenario, although \citet{plotnikova_chemical_2024} showed that multiple sequences can arise due to selection effects, and \citet{chen_recent_2025} showed that radial migration can also lead to such patterns. There are hints of multiple sequences in Figure \ref{fig:yield-outflow}, where a break in the distribution at ages of $\sim5\Gyr$ coincides with a $\sim0.2\dex$ drop in [O/H] and [Fe/H]. The sequences are less pronounced in the \citet{leung_variational_2023} catalog than reported by \citet{anders_spectroscopic_2023}, likely related to a systematic offset between the two catalogs around the same age \citep[see Figure 6 of][]{anders_spectroscopic_2023}. On the other hand, the [C/N] age--metallicity relation of \citet{roberts_cn_2025} is even flatter and does not appear to have multiple sequences (Figure \ref{fig:compare-age-catalogs}). Even if these sequences are real, we find it difficult to explain them using the two-infall scenario\textemdash the observed dilution is much smaller than predicted by any of our models, and the timing of the split between the sequences is too late for the birth of the thin disk, although it is interestingly close to the first pericentric passage of the Sagittarius dwarf galaxy \citep{ruiz-lara_recurrent_2020}.}
\todo{Mention APOGEE selection function?}
% Anders+ use APOGEE clump stars. Their split is 5-6 Gyr ago
% Xiang & Rix use Gaia subgiants. Their split is 8 Gyr ago
% Sahlholdt+ use GALAH turnoff stars. Their split is 7-9 Gyr ago.

\todo{Even if the Anders et al. (2023) sequences are right, it still doesn't look like two-infall. The Anders AMR reaches Solar-ish metallicity 8 Gyr ago, then drops to $\mathFeH\approx-0.2$ by 4 Gyr ago. A 0.2 dex dilution over 4 Gyr is {\it far} weaker and more gradual than predicted by any of our two-infall models.}

@@

The parameters of the SFE hiatus in Figure \ref{fig:onezone-sfe-hiatus} were chosen to match the APOGEE stellar [O/Fe] distribution as closely as possible. As such, we start the hiatus at $t=1.4\Gyr$ to match the high-$\alpha$ locus, even though this is $\sim3\Gyr$ before the median age of the thick disk measured by\citet{pinsonneault_apokasc-3_2025}. This hiatus model illustrates an alternative scenario for reproducing the $\alpha$-bimodality, but a full multi-zone treatment is beyond the scope of this work.

>>>>

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{\figpath{stellar_abundance_evolution.pdf}}
    \caption{Stellar age--abundance relations predicted by multi-zone models at the \yZ{1} yield scale (see Table \ref{tab:yields}). Each point represents a stellar population drawn from the Solar neighborhood near the midplane ($7\leq R_{\rm gal}< 9\kpc$, $|z| < 0.5\kpc$) color-coded by its birth radius. In this and subsequent figures, a Gaussian scatter is applied to each point according to the median age and abundance uncertainties in Table \ref{tab:uncertainties}. For visual clarity, we plot only a random mass-weighted sample of \num{10000} points in each panel. The dashed curve plots the predicted ISM abundance at $R_{\rm gal}=8\kpc$, \added{and the solid black curve plots the rolling median stellar abundance. The red curve plots the rolling median abundance of the APOGEE sample}, and the shaded regions are the 16th--84th percentile ranges. Each column shows results from a different multi-zone model: {\bf (a)} our fiducial model, {\tt yZ1-fiducial}; {\bf (b)} a model with stronger radial migration, {\yZ1-migration}; {\bf (c)} a model with a higher local thick-to-thin disk ratio, {\yZ1-diskratio}; and {\bf (d)} a model with pre-enriched gas infall, {\tt yZ1-preenrich} (see Table \ref{tab:models} for details).}
    \label{fig:abundance-evolution-params}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics[width=\onecolumn]{\figpath{ofe_feh_density.pdf}}
    \caption{The density of stars in the [O/Fe]--[Fe/H] plane predicted by multi-zone models with (a) $y/Z_\odot=1$ and (b) $y/Z_\odot=2$. The curves plot the ISM abundance at the Solar annulus over time, and the alternating black and white segments mark time intervals of 1~Gyr. The model output has been re-sampled to match the APOGEE stellar $|z|$ distribution.
    Stars in \added{both} panels are restricted to the region defined by $7\leq R_{\rm gal}< 9\kpc$ and $|z|<2\kpc$.}
    \label{fig:ofe-feh-density}
\end{figure}

@@

\added{A few recent studies report the discovery of multiple disconnected sequences in the local age--metallicity relation \citep[e.g.,][]{sahlholdt_characterizing_2022,xiang_time-resolved_2022,anders_spectroscopic_2023}. \citet{nissen_high-precision_2020} argued that the existence of multiple sequences is evidence for the two-infall scenario. However, \citet{plotnikova_chemical_2024} showed that multiple sequences can arise due to selection effects, and the APOGEE selection function in particular affects the observed stellar age distribution \citep{imig_galactic_2025}. Furthermore, \citet{chen_recent_2025} demonstrated that multiple sequences in the local AMR can arise naturally due to radial migration, even in the absence of major accretion or starburst events. 
% Anders+ use APOGEE clump stars. Their split is 5-6 Gyr ago
% Xiang & Rix use Gaia subgiants. Their split is 8 Gyr ago
% Sahlholdt+ use GALAH turnoff stars. Their split is 7-9 Gyr ago.

Figure \ref{fig:yield-outflow} shows hints of multiple sequences in the \citet{leung_variational_2023} AMR, specifically where an under-density of $\sim5\Gyr$ old stars coincides with a $\sim0.2\dex$ drop in [O/H] and [Fe/H]. This is less pronounced, but otherwise similar to, the sequences found by \citet{anders_spectroscopic_2023}. We note a systematic offset between the two catalogs around the same age \citep[see Figure 6 of][]{anders_spectroscopic_2023}. On the other hand, the [C/N] AMR of \citet{roberts_cn_2025} is flatter and does not appear to have multiple sequences (Figure \ref{fig:compare-age-catalogs}), although the larger age uncertainties may wash out such patterns.

Even if these AMR sequences are real, the two-infall scenario does not offer a natural explanation. The difference in metallicity between the sequences in our sample is $\sim0.2\dex$, smaller than the dilution predicted by any of our models. Additionally, the transition between the sequences occurs $\sim5\Gyr$ ago \citep[matching][]{anders_spectroscopic_2023}, much younger than the oldest thin disk stars \citep{pinsonneault_apokasc-3_2025}. Interestingly, the transition does align with the first pericentric passage of the Sagittarius dwarf galaxy \citep{ruiz-lara_recurrent_2020}.}

@@

The parameters of the SFE hiatus in Figure \ref{fig:onezone-sfe-hiatus} were chosen to match the APOGEE stellar [O/Fe] distribution as closely as possible. As such, we start the hiatus at $t=1.4\Gyr$ to match the high-$\alpha$ locus, even though this is $\sim3\Gyr$ before the median age of the thick disk measured by \citet{pinsonneault_apokasc-3_2025}. This hiatus model illustrates an alternative scenario for reproducing the $\alpha$-bimodality, but a full multi-zone treatment is beyond the scope of this work.
